---
title: "data_visualisation"
output:
  html_document: default
  pdf_document: default
---

# Data Visualisation
```{r}
library(tidyverse)
mpg
```

## Creating a plot
```{r cache=TRUE}
ggplot(data = mpg) +
    geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

## Graphing template
```{r eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## Facets
 Particularly useful for categorical variables
```{r cache=TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
With two varaibles
```{r cache=TRUE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```
 You could set the shape of a point, but you couldn’t set the “shape” of a line. On the other hand, you could set the linetype of a line. 
```{r cache=TRUE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```
To display multiple geoms in the same plot, add multiple geom functions to `ggplot()`:
```{r cache=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
 You can avoid this type of repetition by passing a set of mappings to `ggplot()`
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```
 If you place mappings in a geom function, ggplot2 will treat them as local mappings for the layer.
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```
 
You can use the same idea to specify different data for each layer. Here, our smooth line displays just a subset of the mpg dataset, the subcompact cars.
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy))+
  geom_point(shape = 20, size = 6, aes(color = drv))+
  geom_smooth(size = 2,se = FALSE)
```

### Exercise
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x= displ, y = hwy))+
  geom_point(mapping = aes(fill = drv), shape = 21, size = 3, stroke = 3, color = "white")

```

## Statistical transformations
```{r cache=TRUE}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut))
```

- bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin.

- smoothers fit a model to your data and then plot predictions from the model.
 
- boxplots compute a robust summary of the distribution and then display a specially formatted box.

The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation. The figure below describes how this process works with `geom_bar()`.

`geom_bar()` uses `stat_count()`

You can generally use geoms and stats interchangeably. For example, you can recreate the previous plot using `stat_count()` instead of `geom_bar()`:
```{r cache=TRUE}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

### There are three reasons you might need to use a stat explicitly:

1. You might want to override the default stat, I change the stat of geom_bar() from count (the default) to identity.
```{r cache=TRUE}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

2. You might want to override the default mapping from transformed variables to aesthetics.
```{r cache=TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

> To find the variables computed by the stat, look for the help section titled “computed variables”.

3. You might want to draw greater attention to the statistical transformation in your code.`stat_summary()`, summarises the y values for each unique x value, to draw attention to the summary that you’re computing:
```{r cache=TRUE}
ggplot(data = diamonds , mapping = aes(x = cut, y = depth)) + 
  stat_summary(
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )
```

### Exercise:

1. What is the default geom associated with `stat_summary()`?
```{r cache=TRUE}
ggplot(data = diamonds, mapping = aes(x = cut, y = depth))+
  geom_pointrange(
    stat = "summary",
      fun.min = min,
      fun.max = max,
      fun = median
    )
```

2. What does `geom_col` do? geom_col represente the values of data with `stat_identity` but `geom_bar`represente the data with `stat_count`. And both are used to plot barcharts.

3. list of pairs geom and stat

      GEOM       |       STAT
-----------------|------------------
geom_bar()       |stat_count()
geom_bin2d()     |stat_bin_2d()
geom_boxplot()   |stat_boxplot()
geom_contour()   |stat_contour()
geom_count()     |stat_sum()
geom_density()   |stat_density()
geom_density_2d()|stat_density_2d()
geom_hex()       |stat_bin_hex()
geom_freqpoly()  |stat_bin()
geom_histogram() |stat_bin()
geom_qq()        |stat_qq()
geom_qq_line()   |stat_qq_line()
geom_quantile()  |stat_quantile()
geom_smooth()    |stat_smooth()
geom_violin()    |stat_ydensity()
geom_sf()        |stat_sf()



4. `stat_smooth` compute this variables:
* `y` predicted value
* `ymin` lower pointwise confidence intrval around the mean
* `ymax` upper pointwise confidence interval around the mean
* `se` standard error

5. In our proportion bar chart, we need to set `group = 1`. Why?
```{r cache=TRUE}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop..))

ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

If group = 1 is not included, then all the bars in the plot will have the same height, a height of 1. The function geom_bar() assumes that the groups are equal to the x values since the stat computes the counts within the group.
The problem with these two plots is that the proportions are calculated within the groups.

```{r cache=TRUE}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```

With the fill aesthetic, the heights of the bars need to be normalized.
```{r cache=TRUE}
ggplot(data = diamonds) +
  geom_bar(aes(x = cut, y = ..count.. / sum(..count..), fill = color))
```

## Position adjustments

The stacking is performed automatically by the **position adjustment** specified by the `position` argument. 
```{r cache=TRUE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

If you don’t want a stacked bar chart, you can use one of three other options: `"identity"`, `"dodge"` or `"fill"`.

**position = "identity"** will place each object exactly where it falls in the context of the graph.this not very useful for bars, because it overlaps them.
```{r cache=TRUE}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(alpha = 1/5, position = "identity")

ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) +
  geom_bar(fill = NA, position = "identity")
```

> The identity position adjustment is more useful for 2d geoms, like points, where it is the default.

**position = "fill"** works like stracking, but makes each of stracked bars the same height. This makes it easier to compare proportions across groups.

```{r cache=TRUE}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

**position = "dodge"** places overlapping objects directly beside one another. This makes it easier to compare individul values.

```{r cache=TRUE}
ggplot(data = diamonds)+
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

> There’s one other type of adjustment that’s not useful for bar charts, but it can be very useful for scatterplots.

The values of hwy and displ are rounded so the points appear on a grid and many points overlap each other. This problem is known as overplotting.
```{r cache=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

You can avoid this gridding by setting the position adjustment to “jitter”. `position = "jitter"` adds a small amount of random noise to each point.
```{r cache=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

**ggplot2 comes with a shorthand for `geom_point(position = "jitter")`: `geom_jitter()`.**
To learn more about a position adjustment, look up the help page associated with each adjustment: ?position_dodge, `?position_fill`, `?position_identity`, `?position_jitter`, and `?position_stack`.

###Exercises:
1. What is the problem with this plot? How could you improve it?
```{r cache=TRUE}
#The problem with this plot is overplotting
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()

#We can improve it by chaging the position to gitter
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
```

2. What parameters to `geom_jitter()` control the amount of jittering?
The parameters that control the behavior of `geom_jitter`is her two arguments:
- width
- height

3.comparaison between `geom_jitter()` and `geom_count`
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
#This method reduces overplotting. However, the reduction in overlapping comes at the cost of slightly changing the x and y values of the points.

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_count()
#The geom_count() geom does not change x and y coordinates of the points. However, if the points are close together and counts are large, the size of some points can itself create overplotting.
```

`geom_count()` is less readable than `geom_jitter()` when adding a third variable as a color aesthetic.
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_jitter()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_count()
```

4. What’s the default position adjustment for `geom_boxplot()`? Create a visualisation of the mpg dataset that demonstrates it.

The default position is dodge2, this default position ovoid overlapping by moving the geom horizontally.
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_boxplot()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy, color = class)) +
  geom_boxplot(position = "identity")
```

## Coordinate systems

The default coordinate system is the Cartesian coordinate system where the x and y positions act independently to determine the location of each point.

### `coord_flip()` 
switches the x and y axes. This is useful (for example), if you want horizontal boxplots. It’s also useful for long labels: it’s hard to get them to fit without overlapping on the x-axis.

```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()

ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

### `coord_quickmap()` 
sets the aspect ratio correctly for maps. This is very important if you’re plotting spatial data with ggplot2
```{r cache=TRUE}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()
```

### `coord_polar()`
uses polar coordinates. Polar coordinates reveal an interesting connection between a bar chart and a Coxcomb chart.

```{r cache=TRUE}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```

### Exercises

1. Turn a stacked bar chart into a pie chart using `coord_polar()`.
```{r cache=TRUE}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
geom_bar()
```

 add `coord_polar(theta="y")` to create pie chart.
```{r cache=TRUE}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
```

 If `coord_polar()` is specified without `theta = "y"`, then the resulting plot is called a bulls-eye chart.
```{r cache=TRUE}
ggplot(mpg, aes(x = factor(1), fill = drv)) +
  geom_bar(width = 1) +
  coord_polar()
```

2. What does `labs()` do? Read the documentation.
```{r  eval= FALSE, cache=TRUE}
?labs
```
labs add labels to the plot.

3. What’s the difference between `coord_quickmap()` and `coord_map()`?


coord_map projects a portion of the earth, which is approximately spherical, onto a flat 2D plane using any projection defined by the mapproj package. Map projections do not, in general, preserve straight lines, so this requires considerable computation. coord_quickmap is a quick approximation that does preserve straight lines. It works best for smaller areas closer to the equator.  

--- From the help page `?coord_map`  

4. What does the plot below tell you about the relationship between city and highway mpg? Why is `coord_fixed()` important? What does `geom_abline()` do?
```{r cache=TRUE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```

`coord_fixed` gives us a real slope of the line that show the relation between the variables.

The function coord_fixed() ensures that the line produced by geom_abline() is at a 45-degree angle. A 45-degree line makes it easy to compare the highway and city mileage to the case in which city and highway MPG were equal.

 The function ggthemes::bank_slopes() will calculate the optimal aspect ratio to bank slopes to 45-degrees.
 
## The layered grammar of graphics

```{r eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```


# Workflow: basics

## Coding basics

All R statements where you create objects, assignment statements, have the same form:

```{r eval=FALSE}
object_name <- value
#When reading that code say “object name gets value” in your head.
```

## Calling functions

```{r cache=TRUE}
y <- seq(1, 10, length.out = 5)

(y <- seq(1, 10, length.out = 5))
```
This common action can be shortened by surrounding the assignment with parentheses, which causes assignment and “print to screen” to happen.

